### 2010-12-23 Documentation/atropos/atropos.tex
ENH: adding Jue Wu results, some discussion, fixes to number of labels.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@709 fd2da0a4-d968-48df-9cb5-d8c496287cbf

\begin{center}
{\Large \bf An Open Source Framework for Multivariate $n$-Tissue
  Segmentation and Evaluation Using Open Data Sets} \\
\vspace*{0.5cm}
{\normalsize Brian B.~Avants$^{1*}$,  Nicholas J.~Tustison$^2$%
\symbolfootnote[1]{
  The first two authors contributed equally to this work.
}, 
Philip A.~Cook$^1$, Jue Woo$^1$, and James C.~Gee$^1$} \\
\begin{singlespace} 
{\scriptsize  $^1$ Penn Image Computing and Science Laboratory, University of Pennsylvania, Philadelphia, Pennsylvania,  USA.\\
  $^2$ Department of Radiology, University of Virginia, Charlottesville, Virginia, USA}
\end{singlespace}
\end{center}

@@

\begin{abstract} 
%Neuroanatomical coordinate systems are essential for the
%interpretation of structural and functional imaging studies.
%However, manual delineation of the sulco-gyral complex is time
%consuming and prone to variability given cortical complexity.  This
%work describes an open source, image-based algorithmic approach to
%cortical parcellation which uses training data to propagate
%structural labelings to individual images.  Along with other
%neuroinformatics tools publicly distributed in our ANTs (Advanced Normalization Tools)%
We introduce {\em Atropos}, an ITK-based multivariate $n$-class
Expectation Maximization (EM) segmentation algorithm based on either parametric or non-parametric finite mixture modeling (FMM) of the
class intensities.  Atropos, available in ANTs
\footnote{
http://www.picsl.upenn.edu/ANTs
}, is capable of incorporating spatial prior probability maps,
(sparse) prior label maps and/or Markov Random Field (MRF) modeling.  Atropos has also been efficiently implemented to handle
large quantities of possible labelings (in the experimental section,
we use up to 68 classes) with a minimal memory footprint.  This work
describes the technical and implementation aspects of Atropos and
evaluates its performance on two different ground-truth datasets.
First, we use the BrainWeb dataset from Montreal Neurological
Institute to evaluate three tissue segmentation performance via (1)
K-means segmentation without use of template data; (2) MRF segmentation with initialization by prior probability maps
derived from a group template; (3) Prior-based segmentation with use
of spatial prior probability maps derived from a group template.  We
also evaluate Atropos performance by using spatial priors to drive a 68 class EM segmentation
problem defined by the Hammers atlas from University College London.  These evaluation studies, combined with
illustrative examples that excercise Atropos options, demonstrate both
performance and wide applicability of this new platform-independent 
open source segmentation tool.
%Component (3) corresponds to the most
%realistic clinical case, while component (2) eliminates the confound
%of defining the cortex itself and focuses only on the parcellation
%problem.  Consistent with the open source dissemination of ANTs, we
%make publicly available all scripts and code from the evaluation.
\end{abstract}

@@

The Atropos algorithm is multi-platform, free, stand-alone and able to
manage a wide range of scalar and vector image datatypes.  To highlight the
value of this open-source contribution, we performed a search of
software attributes on NITRC and found that as of November 2010 no
stand-alone EM methods are currently listed.
Atropos's memory efficiency enables EM-segmentation of many classes at
once, a capability that is due to a sparse internal representation of
prior probability images that minimizes the need for random access
memory.  We evaluate EM-based classification of $1 mm^3$ T1 MR
neuroimages into 68 neuronatomical regions to illustrate the practical
value of the low-memory implementation within this paper.  Atropos
also allows multivariate image features to drive the data term, relevant when more than one view of anatomy aids
segmentation, as in neonatal brain tissue classification
\cite{Weisenfeld2009}.  Although Atropos may be applied to
multivariate data from arbitrary modalities, we limit our evaluation
to tissue-classification in T1 neuroimaging in part due to the
abundance of ``gold-standard'' data for this modality.  Consistent
with our advocacy of open science (not to mention the facilitation of
analysis due to accessibility) we also only use publicly available
data sets.  For this reason, all results in this paper are
reproducible with the caveat that users may require some guidance from
the authors or other users in the community.

@@

\section{Evaluation}
Atropos~encodes a family of segmentation techniques that may be
instantiated for different applications but here we evaluate only two
of the many possibilites. First, we perform an evaluation on the
BrainWeb dataset using both the standard T1 image with multiple bias
and noise levels and also the BrainWeb20 data
\citep{Aubert-Broche2006a,Battaglini2008}.  In combination, this data
allows one to vary not only noise and bias but also the underlying
anatomy.  Second, we evaluate the use of Atropos in improving whole
brain parcellation and excercise its ability to efficiently solve {\em
many class} expectation maximization problem.  We choose this
evaluation problem in part to illustrate the flexibility of Atropos
and also the benefits of the novel, efficient implementation that
allows many-class problems to be solved with low memory usage ($<$2GB
for a 68 class model on 1mm$^3$ brain data).

\subsection{BrainWeb Evaluation}
\label{sec:bweb} The BrainWeb data is freely available at
\url{http://mouldy.bic.mni.mcgill.ca/brainweb/}.  We employ both the
individual subject data and the BrainWeb20 data in this evaluation.
\subsubsection{Single-Subject Evaluation} We use the single-subject
data with 3\% noise and bias 0, 20 and 40 \% RF inhomogeneity.  We
study the effect of the MRF prior term and initialization strategy on
the Dice overlap between ground truth and the segmentation result for
each tissue.  We test both K-means and prior label image
initialization with MRF $\beta \in \{ 0.00 , 0.05 , 0.10 , 0.15 , 0.20
, 0.25 , 0.30 \}$ at each bias field.  We also feed the white matter
probability map derived from K-means into N4 to guide the bias
correction.  Segmentation is then repeated, with the same parameters,
but with the N4 corrected image as input.  The resulting algorithm is
similar to those methods that fix segmentation parameters while
estimating bias and fix bias while estimating segmentation parameters.
Thus, with this simple evaluation, we are able to compare the impact
of bias on our N4 and Atropos algorithm and also the validity of our
prior label image initialization.  The prior label image manual
initialization was based on a crude ITK-SNAP labeling of a small
set of voxels for each tissue.  Results, in terms of
Dice overlap, are shown in figure~\ref{fig:bweb1}.
\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF0.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF20.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF40.pdf}\\
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF0.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF20.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF40.pdf}\\
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF0.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF20.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF40.pdf}
\end{tabular}
\caption{\baselineskip 12pt \small Brainweb single-subject results for
  each tissue.  The results show that the N4 bias correction
  algorithm, combined with Atropos, results in a minimal effect of
  bias, even at the 40\% level.  The optimal $\beta$ for the MRF term
  appears to be between $0.1$ and $0.2$.  The prior label image
  initialization is competitive with but inferior to K-means.}
\label{fig:bweb1}
\end{center}
\end{figure}
\subsubsection{20 Subject Evaluation}

\subsection{The Hammers Dataset Evaluation}
\label{sec:hammers}
We evaluate the ability to improve multi-template
labeling results by converting the group labels to probability maps
and using them to drive many-class EM segmentation.  The procedure we
use here is similar to that described in \citep{Avants2010a}.  The
ground truth labels cover 68 classes and much of the brain.  Some
unlabeled regions remain which we assign to label 69 such that all
brain parenchyma contains a unique label.  The initialization of our
evaluation applies the script {\verb ants_multitemplate_labeling.sh }
(available in the ANTs toolkit) to the 19 Hammers evaluation
datasets located at \url{http://www.brain-development.org/} \citep{Hammers2003,Heckemann2006}.
These initial majority voting results are competitive with prior work
\citep{Heckemann2006,Heckemann2010} and serve as a baseline against
which we compare.

@@

\section{Discussion}

{\textcolor{red}{ Elaborate after including the results.} }

A more general advantage which extends beyond the scope of the experimental 
evaluation section of this paper is the flexibility of the Atropos algorithm.   This includes 
not only $n$-tissue segmentation and dense volumetric cortical parcellation,
as reported in this work, but Atropos is also used in conjunction with our ANTs registration tools for robust
brain extraction which has reported good performance in comparison with other
popular, publicly available brain extraction tools \citep{Avants2010a}.  An additional successful
application area is that of ventilation-based segmentation of hyperpolarized helium-3 MRI
\citep{Tustison2010a}.

%\paragraph{Acknowledgments}
%{This work was supported in part by NIH ... }

>>>>

\begin{center}
{\Large \bf An Open Source Framework for Multivariate $n$-Tissue
  Segmentation and Evaluation Using Open Data Sets} \\
\vspace*{0.5cm}
{\normalsize Brian B.~Avants$^{1*}$,  Nicholas J.~Tustison$^2$%
\symbolfootnote[1]{
  The first two authors contributed equally to this work.
}, 
Jue Wu$^1$, Philip A.~Cook$^1$, and James C.~Gee$^1$} \\
\begin{singlespace} 
{\scriptsize  $^1$ Penn Image Computing and Science Laboratory, University of Pennsylvania, Philadelphia, Pennsylvania,  USA.\\
  $^2$ Department of Radiology, University of Virginia, Charlottesville, Virginia, USA}
\end{singlespace}
\end{center}

@@

\begin{abstract} 
%Neuroanatomical coordinate systems are essential for the
%interpretation of structural and functional imaging studies.
%However, manual delineation of the sulco-gyral complex is time
%consuming and prone to variability given cortical complexity.  This
%work describes an open source, image-based algorithmic approach to
%cortical parcellation which uses training data to propagate
%structural labelings to individual images.  Along with other
%neuroinformatics tools publicly distributed in our ANTs (Advanced Normalization Tools)%
We introduce {\em Atropos}, an ITK-based multivariate $n$-class
Expectation Maximization (EM) segmentation algorithm based on either parametric or non-parametric finite mixture modeling (FMM) of the
class intensities.  Atropos, available in ANTs
\footnote{
http://www.picsl.upenn.edu/ANTs
}, is capable of incorporating spatial prior probability maps,
(sparse) prior label maps and/or Markov Random Field (MRF) modeling.  Atropos has also been efficiently implemented to handle
large quantities of possible labelings (in the experimental section,
we use up to 69 classes) with a minimal memory footprint.  This work
describes the technical and implementation aspects of Atropos and
evaluates its performance on two different ground-truth datasets.
First, we use the BrainWeb dataset from Montreal Neurological
Institute to evaluate three tissue segmentation performance via (1)
K-means segmentation without use of template data; (2) MRF segmentation with initialization by prior probability maps
derived from a group template; (3) Prior-based segmentation with use
of spatial prior probability maps derived from a group template.  We
also evaluate Atropos performance by using spatial priors to drive a 69 class EM segmentation
problem defined by the Hammers atlas from University College London.  These evaluation studies, combined with
illustrative examples that excercise Atropos options, demonstrate both
performance and wide applicability of this new platform-independent 
open source segmentation tool.
%Component (3) corresponds to the most
%realistic clinical case, while component (2) eliminates the confound
%of defining the cortex itself and focuses only on the parcellation
%problem.  Consistent with the open source dissemination of ANTs, we
%make publicly available all scripts and code from the evaluation.
\end{abstract}

@@

The Atropos algorithm is multi-platform, free, stand-alone and able to
manage a wide range of scalar and vector image datatypes.  To highlight the
value of this open-source contribution, we performed a search of
software attributes on NITRC and found that as of November 2010 no
stand-alone EM methods are currently listed.
Atropos's memory efficiency enables EM-segmentation of many classes at
once, a capability that is due to a sparse internal representation of
prior probability images that minimizes the need for random access
memory.  We evaluate EM-based classification of $1 mm^3$ T1 MR
neuroimages into 69 neuronatomical regions to illustrate the practical
value of the low-memory implementation within this paper.  Atropos
also allows multivariate image features to drive the data term, relevant when more than one view of anatomy aids
segmentation, as in neonatal brain tissue classification
\cite{Weisenfeld2009}.  Although Atropos may be applied to
multivariate data from arbitrary modalities, we limit our evaluation
to tissue-classification in T1 neuroimaging in part due to the
abundance of ``gold-standard'' data for this modality.  Consistent
with our advocacy of open science (not to mention the facilitation of
analysis due to accessibility) we also only use publicly available
data sets.  For this reason, all results in this paper are
reproducible with the caveat that users may require some guidance from
the authors or other users in the community.

@@

\section{Evaluation}
Atropos~encodes a family of segmentation techniques that may be
instantiated for different applications but here we evaluate only two
of the many possibilites. First, we perform an evaluation on the
BrainWeb dataset using both the standard T1 image with multiple bias
and noise levels and also the BrainWeb20 data
\citep{Aubert-Broche2006a,Battaglini2008}.  In combination, this data
allows one to vary not only noise and bias but also the underlying
anatomy.  Second, we evaluate the use of Atropos in improving whole
brain parcellation and excercise its ability to efficiently solve {\em
many class} expectation maximization problem.  We choose this
evaluation problem in part to illustrate the flexibility of Atropos
and also the benefits of the novel, efficient implementation that
allows many-class problems to be solved with low memory usage ($<$2GB
for a 69 class model on 1mm$^3$ brain data).

\subsection{BrainWeb Evaluation}
\label{sec:bweb} The BrainWeb data is freely available at
\url{http://mouldy.bic.mni.mcgill.ca/brainweb/}.  We employ both the
individual subject data and the BrainWeb20 data in this evaluation.
\subsubsection{Single-Subject Evaluation} We use the single-subject
data with 3\% noise and bias 0, 20 and 40 \% RF inhomogeneity.  We
study the effect of the MRF prior term and initialization strategy on
the Dice overlap between ground truth and the segmentation result for
each tissue.  We test both K-means and prior label image
initialization with MRF $\beta \in \{ 0.00 , 0.05 , 0.10 , 0.15 , 0.20
, 0.25 , 0.30 \}$ at each bias field.  We also feed the white matter
probability map derived from K-means into N4 to guide the bias
correction.  Segmentation is then repeated, with the same parameters,
but with the N4 corrected image as input.  The resulting algorithm is
similar to those methods that fix segmentation parameters while
estimating bias and fix bias while estimating segmentation parameters.
Thus, with this simple evaluation, we are able to compare the impact
of bias on our N4 and Atropos algorithm and also the validity of our
prior label image initialization.  The prior label image manual
initialization was based on a crude ITK-SNAP labeling of a small
set of voxels for each tissue.  Results, in terms of
Dice overlap, are shown in figure~\ref{fig:bweb1}.
\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF0.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF20.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF40.pdf}\\
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF0.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF20.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF40.pdf}\\
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF0.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF20.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF40.pdf}
\end{tabular}
\caption{\baselineskip 12pt \small Brainweb single-subject results for
  each tissue.  The results show that the N4 bias correction
  algorithm, combined with Atropos, results in a minimal effect of
  bias, even at the 40\% level.  The optimal $\beta$ for the MRF term
  appears to be between $0.1$ and $0.2$.  The prior label image
  initialization is competitive with but inferior to K-means.}
\label{fig:bweb1}
\end{center}
\end{figure}
\subsubsection{20 Subject Evaluation} The single-subject brainweb
study in the previous section tested the basic Atropos options and the
benefit of N4 for segmentation in the presence of bias.  The 20
subject brainweb data allows us to use 2-fold cross-validation to test
our ability to segment different individuals reliably.  In this study,
we divide the 20 subjects equally into training and testing groups.
We then exploit the ground-truth labeling of the training data to
build both a group template \citep{Avants2010} and also prior
probability maps for each of the three major tissues in the cerebrum.
Each prior probability map is gained by deforming the ground truth
labels from each of the 10 training subjects to their template and
averaging component by component.  We then deform the template---and
priors---to the ten testing subjects and run Atropos with not only
{\ttfamily KMeans[3]} initialization but also {\ttfamily
  PriorProbabilityMap[3,priors\%02d.nii.gz,$w$] } where $w \in
\{0.0,0.5\}$.  We then switch the roles of testing and training sets to
gain 3-tissue segmentation for each of the twenty subjects.  
When $w=0$, the priors are only used in initializing
the model parameters but not during subsequent EM iterations.  
When $w=1$, the priors are only maintained in the product with the
likelihood during all EM iterations.  Results, in terms of bar
plots for Dice overlap mean and standard deviation, are shown in figure~\ref{fig:bweb20}.
\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[width=2.2in]{Figures/bw20_kmeans.png}
\includegraphics[width=2.2in]{Figures/bw20_priorIni.png}
\includegraphics[width=2.2in]{Figures/bw20_priorSeg.png}
\end{tabular}
\caption{\baselineskip 12pt \small Brainweb 20-subject results for
  each tissue as a function of MRF-$\beta$ parameter where MRF-$\beta$
  is in $\{0,0.05,0.1,0.15,0.2\}$ and increases left to right.  The results show that the PriorProbabilityMaps with
  $w=0.5$ (far right) gives the best performance for all tissues.}
\label{fig:bweb20}
\end{center}
\end{figure}

\subsection{The Hammers Dataset Evaluation}
\label{sec:hammers}
We evaluate the ability to improve multi-template
labeling results by converting the group labels to probability maps
and using them to drive many-class EM segmentation.  The procedure we
use here is similar to that described in \citep{Avants2010a}.  The
ground truth labels cover 69 classes and much of the brain.  Some
unlabeled regions remain which we assign to label 69 such that all
brain parenchyma contains a unique label.  The initialization of our
evaluation applies the script {\verb ants_multitemplate_labeling.sh }
(available in the ANTs toolkit) to the 19 Hammers evaluation
datasets located at \url{http://www.brain-development.org/} \citep{Hammers2003,Heckemann2006}.
These initial majority voting results are competitive with prior work
\citep{Heckemann2006,Heckemann2010} and serve as a baseline against
which we compare.

@@

\subsection{Reproducibility of this Evaluation}
The brainweb data is freely available.  We used single-subject
brainweb data as is.  The 20 subject data, however, required excluding
non-cerebrum tissue classes.  The Hammers data was also used as is.
The scripts for the single-subject brainweb study is available in
supplementary material.  The initialization procedure for the brainweb 20 and
the Hammers evaluation data is based on freely available scripts
included with the ANTs toolkit.  {\bf \textcolor{red}{ FIXME}}.

\section{Discussion}
We introduced the Atropos algorithm, the theory and implementation
details and documented its performance in a variety of use cases.  We
also showed evidence that the openly available N4 bias correction
method can easily used with Atropos to improve segmentation.
Furthermore, we used multiple subject brainweb data to build dataset
specific priors that provided the most consistent segmentation
performance across tissues.  Finally, we used majority voting to
initialize an Atropos EM solution to a 69 class brain parcellation
problem.  Significant improvements were gained in multiple brain
regions, in particular in temporal lobe cortex, the hippocampi and
amygdalae and the lateral ventricles.  This work, in summary, proves
the applicability of Atropos in both basic and extended use cases.

\subsection{Performance on Single-Subject Brainweb Data}
How does Atropos performance compare to other methods that have been
evaluated on the basic brainweb data?   {\bf {\textcolor{red}{ FIXME}}}.

\subsection{Performance on Twenty-Subject Brainweb Data}
How does Atropos performance compare to other methods that have been
evaluated on the brainweb twenty data?   {\bf {\textcolor{red}{ FIXME}}}.

\subsection{Performance on Hammers Data}
Our prior work, \citep{Avants2010b}, showed that the majority vote
initialization provided to Atropos by ANTs template mapping is
competitive with \cite{Heckemann2006}.  The Atropos EM extension of
improved these results further.  However, the Atropos EM segmentation
performed significantly worse.  This is not surprising, in that
Atropos EM assumes that signal from the likelihood and MRF term is
valuable in improving the segmentation.  This assumption held for
amygdala and lateral ventricles among other areas.  However, in
pallidum and corpus callosum (the most significant areas with loss of
performance), this is not true.  We believe the explanation is that
the intensity varies within these structures and that a more complex
intensity model (or finer parcellation) would be needed here.  An
alternative solution would be to use boundary conditions for these
structures, as in the PriorLabelImage Atropos option.

\subsection{Other Thoughts} {\bf {\textcolor{red}{ FIXME}}}
A more general advantage which extends beyond the scope of the experimental 
evaluation section of this paper is the flexibility of the Atropos algorithm.   This includes 
not only $n$-tissue segmentation and dense volumetric cortical parcellation,
as reported in this work, but Atropos is also used in conjunction with our ANTs registration tools for robust
brain extraction which has reported good performance in comparison with other
popular, publicly available brain extraction tools \citep{Avants2010a}.  An additional successful
application area is that of ventilation-based segmentation of hyperpolarized helium-3 MRI
\citep{Tustison2010a}.

\section{Conclusion}
The Atropos algorithm is freely available to the public.  We release
this code not only to make it available to clinical researchers but
with the hope that other researchers in segmentation will provide
feedback about the implementation decisions that we made.  EM
segmentation is non-trivial and there are numerous design alternatives
available not only in the models selected but also in the ICM coding
strategy, alternatives to ICM and the specific method in which prior
and likelihood are combined.  Due to the flexibility of Atropos, we
also hope that some of its capabilities, though not evaluated here,
are explored by the segmentation or clinical community.

\paragraph{Acknowledgments}
{This work was supported in part by NIH  {\bf {\textcolor{red}{ FIXME}}} }
